---
title: "abs_vs_sim_eye_tracking_analysis"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```


Loads required packages and reads model_predictions.csv, a preprocessed dataset generated from notebooks/empirical_gaze_analysis.ipynb. 

```{r}
library(readr)
library(dplyr)
library(purrr)
library(tidyr)
library(tibble)
library(ggplot2)
model_predictions <- read_csv(".../model_predictions.csv")

scene_to_screen_coordinates <- function(scene_x, scene_y,
                                        scene_x_lim = 800, scene_y_lim = 1000,
                                        screen_x_lim = 1920, screen_y_lim = 1080) {
  converted_x <- screen_x_lim / 2 - scene_x_lim / 2 + scene_x
  converted_y <- screen_y_lim / 2 - scene_y_lim / 2 + scene_y
  c(converted_x, converted_y)
}

blended_model_data <- within(model_predictions, {
  x <- 1920/2 - 800/2 + x
  y <- 1080/2 - 1000/2 + y
})

names(blended_model_data)[names(blended_model_data) == "segment_class"] <- "reasoning_trace"
```

Creates start and end coordinates for each model segment, so abstraction and simulation traces can be treated as line segments rather than isolated points.

```{r}
library(dplyr)

test <- blended_model_data %>%
  group_by(scene_name, sample_idx) %>%
  mutate(
    start_x = if_else(
      reasoning_trace %in% c("abstraction", "simulation"),
      lag(x, n = 1),
      x
    ),
    start_y = if_else(
      reasoning_trace %in% c("abstraction", "simulation"),
      lag(y, n = 1),
      y
    ),
    end_x = x,
    end_y = y
  ) %>%
  ungroup()

```






# Preprocessing: Make simulation segments visible

Splits the first abstraction segment after each simulation point, relabeling the first 20% as simulation. This makes short simulation traces easier to visualize.

```{r}
library(dplyr)
split_frac <- 0.20  # 20%

test <- blended_model_data %>%
  group_by(scene_name, sample_idx) %>%
  mutate(
    row_id     = row_number(),
    prev_trace = lag(reasoning_trace),

    # build segments (prev point -> current point) for the two traces
    start_x = if_else(reasoning_trace %in% c("abstraction", "simulation"), lag(x), x),
    start_y = if_else(reasoning_trace %in% c("abstraction", "simulation"), lag(y), y),
    end_x   = x,
    end_y   = y
  ) %>%
  ungroup() %>%
  mutate(
    # identify "start of abstraction segment right after simulation"
    is_split = reasoning_trace == "abstraction" & prev_trace == "simulation" &
               !is.na(start_x) & !is.na(start_y),

    # point that is 20% along the abstraction segment (from start -> end)
    mid_x = start_x + split_frac * (end_x - start_x),
    mid_y = start_y + split_frac * (end_y - start_y)
  )

# Build the new, split dataset:
test <- bind_rows(
  # rows that are not split
  test %>%
    filter(!is_split) %>%
    mutate(row_id2 = row_id),

  # the new "simulation" piece: start -> mid (20%)
  test %>%
    filter(is_split) %>%
    mutate(
      reasoning_trace = "simulation",
      end_x = mid_x, end_y = mid_y,
      row_id2 = row_id + 0.1
    ),

  # the remaining "abstraction" piece: mid -> end (80%)
  test %>%
    filter(is_split) %>%
    mutate(
      start_x = mid_x, start_y = mid_y,
      row_id2 = row_id + 0.2
    )
) %>%
  arrange(scene_name, sample_idx, row_id2) %>%
  dplyr::select(-row_id2, -prev_trace, -is_split, -mid_x, -mid_y)

```


# Segment-based matching

Instead of matching to every interpolated point, we match human data to each model segment (abstraction or simulation).

For each model segment, we collect human points that satisfy both:

1. Within the segment’s y-range: $y \in \left[\min(\mathrm{start\_y}, \mathrm{end\_y}),\ \max(\mathrm{start\_y}, \mathrm{end\_y})\right]$;

2. Close to the segment: point-to-segment distance $\le \mathrm{dist\_thresh}$ (perpendicular distance $\le 20$).

This avoids repeatedly reusing the same human samples across many nearby model points, while still aligning eye-tracking data to the model’s predicted trace.



# Scene filtering

Reads classified_trial_data.csv, a preprocessed dataset generated from notebooks/empirical_gaze_analysis.ipynb. 


```{r}
classified_trial_data <- read_csv(".../classified_trial_data.csv")
```

Removes scenes where the model trace is physically implausible or shows problematic abstraction patterns.


```{r}
library(dplyr)

drop_scenes <- c(
  "med_nocol_nosp_1",
  "high_yescol_yessp_1",
  "high_nocol_nosp_2",
  "high_nocol_yessp_1",
  "high_yescol_yessp_4",
  "high_nocol_yessp_2",
  "high_yescol_yessp_3",
  "low_yescol_nosp_2"
)

test <- test %>% filter(!scene_name %in% drop_scenes)
classified_trial_data <- classified_trial_data %>% filter(!scene_name %in% drop_scenes)
```

Converts gaze variables to numeric values, computes gaze speed within each subject and scene, and log-transforms pupil size.


```{r}
classified_trial_data <- classified_trial_data %>%
  dplyr::mutate(
    normalized_time = as.numeric(normalized_time),
    x = as.numeric(x),
    y = as.numeric(y)
  ) %>%
  dplyr::group_by(scene_name, subject_id) %>%
  dplyr::arrange(normalized_time, .by_group = TRUE) %>%
  dplyr::mutate(
    gaze_speed = sqrt((x - dplyr::lag(x))^2 + (y - dplyr::lag(y))^2),
    gaze_speed = dplyr::if_else(
      dplyr::row_number() == 1,
      dplyr::lead(gaze_speed),
      gaze_speed
    )
  ) %>%
  dplyr::ungroup()
classified_trial_data$pupil <- log(classified_trial_data$pupil)
```



Splits the human eye-tracking dataset into one dataframe per subject for subject-level analyses.

```{r,results='hide'}
dfs <- classified_trial_data %>%
  dplyr::group_by(subject_id) %>%
  dplyr::group_split(.keep = TRUE)

subject_names <- classified_trial_data %>%
  dplyr::group_by(subject_id) %>%
  dplyr::group_keys() %>%
  dplyr::pull(subject_id)

base::names(dfs) <- subject_names
base::length(dfs)      
base::names(dfs)   
```


Prints the number of Experiment 1 and Experiment 2 scenes available for each participant.

```{r}
for (nm in names(dfs)) {
  scene_names <- unique(dfs[[nm]]$scene_name)

  n_starts_scene <- sum(grepl("^scene", scene_names))
  n_not_starts_scene <- sum(!grepl("^scene", scene_names))

  cat(
    nm, ":",
    n_not_starts_scene, "scenes in Exp 1,",
    n_starts_scene, "scenes in Exp 2\n"
  )
}
```


# Saccade and smooth pursuit probabilities during abstraction vs simulation

Matches human gaze points to nearby abstraction or simulation model segments using y-range filtering and point-to-segment distance. For each matched segment, it counts saccades, smooth pursuits, fixations, and PSOs.

```{r, cache=TRUE}
point_to_segment_dist <- function(px, py, ax, ay, bx, by) {
  # returns distance from point P(px,py) to segment AB(ax,ay)-(bx,by)
  abx <- bx - ax
  aby <- by - ay
  apx <- px - ax
  apy <- py - ay

  denom <- abx^2 + aby^2

  # handle zero-length segments (A == B)
  t <- ifelse(denom == 0, 0, (apx * abx + apy * aby) / denom)
  t <- pmin(1, pmax(0, t))

  cx <- ax + t * abx
  cy <- ay + t * aby

  sqrt((px - cx)^2 + (py - cy)^2)
}

library(dplyr)
library(tibble)

results_prop_by_subject <- list()

dist_thresh <- 20  # perpendicular distance threshold (pixels, same scale as x/y)

n_subjects <- length(dfs)
counter <- 0

for (subject_id in names(dfs)) {

  counter <- counter + 1
  message("[", counter, "/", n_subjects, "] Processing subject: ", subject_id)

  classified_subject <- dfs[[subject_id]]
  results_prop_list <- list()

  for (scene in unique(test$scene_name)) {

    # --- model segments for this scene (abstraction/simulation only)
    segs <- test %>%
      filter(scene_name == scene,
             reasoning_trace %in% c("abstraction", "simulation")) %>%
      dplyr::select(scene_name, reasoning_trace, start_x, start_y, end_x, end_y)

    # --- human points for this scene
    human <- classified_subject %>%
      filter(scene_name == scene) %>%
      dplyr::select(scene_name, x, y, segment_class)

    if (nrow(segs) == 0 || nrow(human) == 0) {
      results_prop_list[[scene]] <- NULL
      next
    }

    rows <- vector("list", nrow(segs))
    keep <- 0

    # loop over model segments
    for (i in seq_len(nrow(segs))) {

      ax <- segs$start_x[i]; ay <- segs$start_y[i]
      bx <- segs$end_x[i];   by <- segs$end_y[i]

      # skip if any endpoints missing/non-finite
      if (!is.finite(ax) || !is.finite(ay) || !is.finite(bx) || !is.finite(by)) next

      # 1) filter by y-range of the segment
      y_lo <- min(ay, by)
      y_hi <- max(ay, by)
      cand <- human %>% filter(y >= y_lo, y <= y_hi)

      if (nrow(cand) == 0) next

      # 2) perpendicular distance to segment
      d <- point_to_segment_dist(cand$x, cand$y, ax, ay, bx, by)
      nearby <- cand[d <= dist_thresh, , drop = FALSE]

      if (nrow(nearby) == 0) next

      keep <- keep + 1
      rows[[keep]] <- tibble(
        scene_name      = scene,
        reasoning_trace = segs$reasoning_trace[i],
        start_x = ax, start_y = ay,
        end_x   = bx, end_y   = by,

        n_within_1      = nrow(nearby),
        n_smooth_pursuit = sum(nearby$segment_class == "smooth_pursuit", na.rm = TRUE),
        n_saccade        = sum(nearby$segment_class == "saccade",        na.rm = TRUE),
        n_fixation       = sum(nearby$segment_class == "fixation",       na.rm = TRUE),
        n_pso            = sum(nearby$segment_class == "pso",            na.rm = TRUE)
      )
    }

    if (keep == 0) {
      results_prop_list[[scene]] <- NULL
      next
    }

    results_prop_list[[scene]] <- bind_rows(rows[seq_len(keep)]) %>%
      mutate(
        prop_smooth   = n_smooth_pursuit / n_within_1,
        prop_saccade  = n_saccade / n_within_1,
        prop_fixation = n_fixation / n_within_1,
        prop_pso      = n_pso / n_within_1
      )
  }

  results_prop_by_subject[[subject_id]] <- results_prop_list
  message("Completed subject: ", subject_id, "\n")
}

```

For each participant, calculates mean saccade and smooth-pursuit proportions during abstraction and simulation.

```{r}
results_summary_by_subject <- list()

for (subject_id in names(results_prop_by_subject)) {

  scene_list <- results_prop_by_subject[[subject_id]]

  # Combine all scenes for this participant
  df_all <- purrr::map_dfr(
    scene_list,
    function(x) {
      if (is.null(x) || nrow(x) == 0) return(NULL)
      x
    }
  )

  # If no valid data, skip
  if (nrow(df_all) == 0) {
    results_summary_by_subject[[subject_id]] <- NULL
    next
  }

  # Keep only abstraction / simulation
  df_all <- df_all %>%
    dplyr::filter(reasoning_trace %in% c("abstraction", "simulation"))

  # If one of the groups is missing, still compute what exists
  summary_df <- df_all %>%
    dplyr::group_by(reasoning_trace) %>%
    dplyr::summarise(
      mean_prop_saccade = mean(prop_saccade, na.rm = TRUE),
      mean_prop_smooth  = mean(prop_smooth,  na.rm = TRUE),
      n_model_point     = dplyr::n(),
      .groups = "drop"
    ) %>%
    # Ensure both rows exist
    dplyr::right_join(
      tibble::tibble(reasoning_trace = c("abstraction", "simulation")),
      by = "reasoning_trace"
    )

  results_summary_by_subject[[subject_id]] <- summary_df
}

```

Creates a wide subject-level table with separate abstraction and simulation columns for saccade and smooth-pursuit proportions.

```{r}
library(dplyr)
library(purrr)
library(tidyr)
library(tibble)

summary_wide_by_subject <- purrr::imap_dfr(
  results_prop_by_subject,
  function(scene_list, subject_id) {

    df_all <- purrr::map_dfr(
      scene_list,
      function(x) {
        if (is.null(x) || nrow(x) == 0) return(NULL)
        x
      }
    )

    if (nrow(df_all) == 0) return(NULL)

    df_all <- df_all %>%
      dplyr::filter(reasoning_trace %in% c("abstraction", "simulation"))

    summary_df <- df_all %>%
      dplyr::group_by(reasoning_trace) %>%
      dplyr::summarise(
        mean_prop_saccade = mean(prop_saccade, na.rm = TRUE),
        mean_prop_smooth  = mean(prop_smooth, na.rm = TRUE),
        .groups = "drop"
      ) %>%
      dplyr::right_join(
        tibble::tibble(reasoning_trace = c("abstraction", "simulation")),
        by = "reasoning_trace"
      ) %>%
      tidyr::pivot_wider(
        names_from = reasoning_trace,
        values_from = c(mean_prop_saccade, mean_prop_smooth)
      ) %>%
      dplyr::mutate(subject_id = subject_id, .before = 1)

    summary_df
  }
)

summary_wide_by_subject
```




Aggregates results across participants and reports average saccade and smooth-pursuit proportions for abstraction versus simulation.

```{r}
group_level_summary <- purrr::map_dfr(
  results_summary_by_subject,
  function(df) {
    if (is.null(df) || nrow(df) == 0) return(NULL)
    df
  },
  .id = "subject_id"
) %>%
  dplyr::group_by(reasoning_trace) %>%
  dplyr::summarise(
    avg_mean_prop_saccade = mean(mean_prop_saccade, na.rm = TRUE),
    avg_mean_prop_smooth  = mean(mean_prop_smooth,  na.rm = TRUE),
    n_subjects            = sum(!is.na(mean_prop_saccade) | !is.na(mean_prop_smooth)),
    .groups = "drop"
  )
knitr::kable(group_level_summary)
```



### Statistical analysis


Combines segment-level results across subjects and scenes, then prepares variables for statistical analysis of saccade versus smooth-pursuit differences.


```{r}
library(dplyr)
library(purrr)

combined_df <- imap_dfr(results_prop_by_subject, function(subject_list, subject_id) {

  imap_dfr(subject_list, function(df, scene_id) {

    if (!is.data.frame(df)) return(NULL)

    df %>%
      select(n_saccade, n_smooth_pursuit, prop_smooth, prop_saccade, reasoning_trace) %>%
      mutate(
        subject_id = subject_id,
        scene_id   = scene_id,
        prop_diff  = prop_saccade - prop_smooth,
        reasoning_trace = factor(
          ifelse(reasoning_trace == "abstraction", 1, 0),
          levels = c(0, 1),
          labels = c("simulation", "abstraction")
        )
      )
  })
})

```


#### Logit difference


Computes the log-odds difference between saccade and smooth pursuit, then fits a mixed-effects model testing whether this difference changes between simulation and abstraction.

The log odds ratio of saccades vs. smooth pursuit:

- logit_diff = 0: equal odds;

- logit_diff > 0: higher odds of saccades;

- logit_diff < 0: higher odds of smooth pursuit.



```{r}
logit <- function(p) log(p / (1 - p))
epsilon <- 1e-5

combined_df <- combined_df %>%
  mutate(
    logit_saccade = logit(pmin(pmax(prop_saccade, epsilon), 1 - epsilon)),
    logit_smooth  = logit(pmin(pmax(prop_smooth,  epsilon), 1 - epsilon)),
    logit_diff    = logit_saccade - logit_smooth
  )

library(lme4)
library(lmerTest)

model_ld <- lmer(
  logit_diff ~ reasoning_trace + (1 + reasoning_trace | subject_id),
  data = combined_df
)

summary(model_ld)
```










Creates a multi-panel figure showing model traces, gaze-density patterns, segment-level logit differences, and group-level movement proportions.

```{r, fig.height=3.5, fig.width=16}
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)
library(gridExtra)
library(grid)
library(png)
library(readr)
library(ggnewscale)

# =========================================================
# (a) Group-level bar plot
# =========================================================
group_level_summary_sem <- purrr::map_dfr(
  results_summary_by_subject,
  function(df) {
    if (is.null(df) || nrow(df) == 0) return(NULL)
    df
  },
  .id = "subject_id"
) %>%
  dplyr::filter(reasoning_trace %in% c("abstraction", "simulation")) %>%
  tidyr::pivot_longer(
    cols = c(mean_prop_saccade, mean_prop_smooth),
    names_to = "movement",
    values_to = "value"
  ) %>%
  dplyr::mutate(
    condition = dplyr::recode(
      reasoning_trace,
      abstraction = "Abstraction",
      simulation  = "Simulation"
    ),
    movement = dplyr::recode(
      movement,
      mean_prop_saccade = "Saccade",
      mean_prop_smooth  = "Smooth pursuit"
    )
  ) %>%
  dplyr::group_by(movement, condition) %>%
  dplyr::summarise(
    mean_value = mean(value, na.rm = TRUE),
    sd_value   = sd(value, na.rm = TRUE),
    n_subjects = sum(!is.na(value)),
    sem_value  = sd_value / sqrt(n_subjects),
    .groups = "drop"
  )

p_bar <- ggplot(
  group_level_summary_sem,
  aes(x = condition, y = mean_value, fill = condition)
) +
  geom_col(width = 0.6) +
  geom_errorbar(
    aes(
      ymin = mean_value - sem_value,
      ymax = mean_value + sem_value
    ),
    width = 0.2
  ) +
  facet_wrap(~ movement, nrow = 1) +
  scale_fill_manual(
    values = c(
      "Abstraction" = "coral",
      "Simulation"  = "lightgreen"
    )
  ) +
  labs(
    x = NULL,
    y = "Average mean proportion (± SEM)",
    fill = NULL,
    title = " ",
    tag = "(d)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(size = 9, angle = 30, hjust = 1),
    axis.text.y = element_text(size = 9),
    axis.title.y = element_text(size = 10),
    strip.text = element_text(size = 10),
    plot.tag = element_text(face = "plain"),
    plot.tag.position = c(0, 1)
  )

# =========================================================
# Scene plots
# =========================================================
scene <- "high_yescol_nosp_1"
scene_image_dir <- ".../intuitive_physics/eyetracking_physics-master/data/scene_images"

xmin <- 560; xmax <- 1360
ymin <- 40;  ymax <- 1040

scene_seg_stats <- purrr::imap_dfr(results_prop_by_subject, function(scene_list, subject_id) {
  df <- scene_list[[scene]]
  if (is.null(df) || nrow(df) == 0) return(NULL)

  df %>%
    dplyr::mutate(
      subject_id = subject_id,
      logit_saccade = logit(pmin(pmax(prop_saccade, epsilon), 1 - epsilon)),
      logit_smooth  = logit(pmin(pmax(prop_smooth,  epsilon), 1 - epsilon)),
      logit_diff    = logit_saccade - logit_smooth
    ) %>%
    dplyr::select(
      scene_name, reasoning_trace,
      start_x, start_y, end_x, end_y,
      logit_diff, subject_id
    )
})

seg_mean_sacc <- scene_seg_stats %>%
  dplyr::filter(!is.na(logit_diff)) %>%
  dplyr::group_by(scene_name, reasoning_trace, start_x, start_y, end_x, end_y) %>%
  dplyr::summarise(
    mean_logit_diff = mean(logit_diff, na.rm = TRUE),
    n_subjects = dplyr::n_distinct(subject_id),
    .groups = "drop"
  )

plot_df1 <- test %>%
  dplyr::filter(
    scene_name == scene,
    reasoning_trace %in% c("abstraction", "simulation")
  ) %>%
  dplyr::select(scene_name, reasoning_trace, start_x, start_y, end_x, end_y) %>%
  dplyr::left_join(
    seg_mean_sacc,
    by = c("scene_name", "reasoning_trace", "start_x", "start_y", "end_x", "end_y")
  )

plot_df2 <- test %>%
  dplyr::filter(
    scene_name == scene,
    reasoning_trace %in% c("abstraction", "simulation")
  )

img_path <- base::file.path(scene_image_dir, paste0(scene, ".png"))
bg_img <- png::readPNG(img_path)

base_scene_plot <- ggplot2::ggplot() +
  ggplot2::annotation_raster(
    bg_img,
    xmin = xmin, xmax = xmax,
    ymin = ymin, ymax = ymax
  ) +
  ggplot2::coord_equal(
    xlim = c(xmin, xmax),
    ylim = c(ymin, ymax),
    expand = FALSE
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(
    aspect.ratio = (ymax - ymin) / (xmax - xmin)
  )

# (b) previous plot 1
lim <- max(abs(plot_df1$mean_logit_diff), na.rm = TRUE)

p1 <- base_scene_plot +
  ggplot2::geom_segment(
    data = plot_df1,
    ggplot2::aes(
      x = start_x, y = 1080 - start_y,
      xend = end_x, yend = 1080 - end_y,
      color = mean_logit_diff
    ),
    linewidth = 0.7, alpha = 0.9, lineend = "round"
  ) +
  ggplot2::scale_color_gradientn(
    colors = c("darkgreen", "lightgreen", "coral"),
    limits = c(0, lim),
    oob = scales::squish,
    name = "Mean logit\nsaccade - smooth",
    guide = ggplot2::guide_colorbar(reverse = FALSE)
  ) +
  ggplot2::labs(
    x = NULL, y = NULL,
    title = " ",
    tag = "(c)"
  ) +
  ggplot2::theme(
    plot.tag = element_text(size = 14, face = "plain"),
    plot.tag.position = c(0, 1),
    plot.margin = margin(0, 0, 0, 0)
  )

# previous plot 2
p2 <- base_scene_plot +
  ggplot2::geom_segment(
    data = plot_df2,
    ggplot2::aes(
      x = start_x, y = 1080 - start_y,
      xend = end_x, yend = 1080 - end_y,
      color = reasoning_trace
    ),
    linewidth = 0.6, alpha = 0.8, lineend = "round"
  ) +
  ggplot2::geom_point(
    data = plot_df2,
    ggplot2::aes(
      x = end_x, y = 1080 - end_y,
      color = reasoning_trace
    ),
    size = 1.2, alpha = 0.9
  ) +
  ggplot2::scale_color_manual(
    values = c(abstraction = "coral", simulation = "lightgreen"),
    breaks = c("abstraction", "simulation"),
    labels = c("Abstraction", "Simulation")
  ) +
  ggplot2::labs(
    x = NULL, y = NULL,
    color = NULL,
    title = NULL,
    tag = "(a)"
  ) +
  ggplot2::guides(
    color = ggplot2::guide_legend(
      override.aes = list(
        linewidth = 2,
        alpha = 1
      )
    )
  ) +
  ggplot2::theme(
    legend.position = "inside",
    legend.position.inside = c(0.03, 0.97),
    legend.justification = c(0, 1),
    legend.background = ggplot2::element_rect(
      fill = scales::alpha("white", 0.7),
      color = NA
    ),
    plot.title = ggplot2::element_blank(),
    plot.tag = element_text(size = 14, face = "plain"),
    plot.tag.position = c(0, 1),
    plot.margin = margin(0, 0, 0, 0)
  )
# =========================================================
# (d) Gaze-density plot
# =========================================================
classified_trial_data <- readr::read_csv(
  ".../classified_trial_data.csv"
)

get_mode_first <- function(x) {
  x <- x[!is.na(x)]
  if (length(x) == 0) return(NA_character_)
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}

mean_segment_gaze_data <- classified_trial_data %>%
  dplyr::mutate(
    x_int = if ("x_int" %in% names(.)) x_int else as.integer(round(x)),
    y_int = if ("y_int" %in% names(.)) y_int else as.integer(round(y))
  ) %>%
  dplyr::group_by(scene_name, normalized_time) %>%
  dplyr::summarise(
    x_int = mean(x_int, na.rm = TRUE),
    y_int = mean(y_int, na.rm = TRUE),
    segment_class = get_mode_first(segment_class),
    pupil = mean(pupil, na.rm = TRUE),
    .groups = "drop"
  )

scene_gaze_data <- mean_segment_gaze_data %>%
  dplyr::filter(
    scene_name == scene,
    segment_class %in% c("saccade", "smooth_pursuit")
  ) %>%
  dplyr::mutate(
    y_plot = 1080 - y_int
  )

legend_df <- tibble::tibble(
  segment_class = c("saccade", "smooth_pursuit"),
  x = c(NA_real_, NA_real_),
  y = c(NA_real_, NA_real_)
)

p3 <- base_scene_plot +
  stat_bin_2d(
    data = scene_gaze_data %>% dplyr::filter(segment_class == "saccade"),
    aes(x = x_int, y = y_plot, fill = after_stat(count)),
    bins = 40,
    alpha = 0.85
  ) +
  scale_fill_gradient(
    low = "#FFD6CC",
    high = "coral",
    na.value = "transparent",
    guide = "none"
  ) +
  ggnewscale::new_scale_fill() +
  stat_bin_2d(
    data = scene_gaze_data %>% dplyr::filter(segment_class == "smooth_pursuit"),
    aes(x = x_int, y = y_plot, fill = after_stat(count)),
    bins = 40,
    alpha = 0.85
  ) +
  scale_fill_gradient(
    low = "#CFECCF",
    high = "green3",
    na.value = "transparent",
    guide = "none"
  ) +
  geom_point(
    data = legend_df,
    aes(x = x, y = y, color = segment_class),
    shape = 15,
    size = 4
  ) +
  scale_color_manual(
    values = c(
      "saccade" = "coral",
      "smooth_pursuit" = "green3"
    ),
    labels = c(
      "saccade" = "Saccade",
      "smooth_pursuit" = "Smooth pursuit"
    ),
    name = NULL
  ) +
  guides(
    color = guide_legend(
      override.aes = list(shape = 15, size = 5, alpha = 1)
    )
  ) +
  labs(
    x = NULL, y = NULL,
    title = " ",
    tag = "(b)"
  ) +
  theme(
    legend.position = "inside",
    legend.position.inside = c(0.03, 0.97),
    legend.justification = c(0, 1),
    legend.background = ggplot2::element_rect(
      fill = scales::alpha("white", 0.7),
      color = NA
    ),
    plot.tag = element_text(size = 14, face = "plain"),
    plot.tag.position = c(0, 1),
    plot.margin = margin(0, 0, 0, 0)
  )
# =========================================================
# Arrange plots
# =========================================================
g_bar <- ggplotGrob(p_bar)
g1 <- ggplotGrob(p2)
g2 <- ggplotGrob(p1)
g3 <- ggplotGrob(p3)

# make scene plots identical in panel/axis/legend layout
common_widths  <- unit.pmax(g1$widths,  g2$widths,  g3$widths)
common_heights <- unit.pmax(g1$heights, g2$heights, g3$heights)

g1$widths  <- common_widths
g2$widths  <- common_widths
g3$widths  <- common_widths

g1$heights <- common_heights
g2$heights <- common_heights
g3$heights <- common_heights


# =========================================================
# Display all four in one row
# =========================================================

blank <- nullGrob()   # or rectGrob(gp = gpar(col = NA, fill = NA))


g_bar_short <- arrangeGrob(
  nullGrob(),
  g_bar,
  nullGrob(),
  ncol = 1,
  heights = c(0.06, 0.88, 0.06)
)

library(grid)
library(gridExtra)

combined_figure <- arrangeGrob(
  blank, g1, g3, g2, blank, g_bar_short,
  ncol = 6,
  widths = c(0.2, 1, 1, 1, 0.1, 0.8)
)

grid.newpage()
grid.draw(combined_figure)
```









#  Log pupil size difference during abstraction vs simulation


Repeats segment-based matching, but this time extracts mean pupil size, overall gaze speed, and saccade-only gaze speed for each matched model segment.

```{r, cache=TRUE}
library(dplyr)
library(tibble)

point_to_segment_dist <- function(px, py, ax, ay, bx, by) {
  abx <- bx - ax
  aby <- by - ay
  apx <- px - ax
  apy <- py - ay

  denom <- abx^2 + aby^2
  t <- ifelse(denom == 0, 0, (apx * abx + apy * aby) / denom)
  t <- pmin(1, pmax(0, t))

  cx <- ax + t * abx
  cy <- ay + t * aby

  sqrt((px - cx)^2 + (py - cy)^2)
}

results_prop_by_subject <- list()

dist_thresh <- 20  # perpendicular distance threshold (px)

n_subjects <- length(dfs)
counter <- 0

for (subject_id in names(dfs)) {

  counter <- counter + 1
  message("[", counter, "/", n_subjects, "] Processing subject: ", subject_id)

  classified_subject <- dfs[[subject_id]]
  results_prop_list <- list()

  for (scene in unique(test$scene_name)) {

    # Model segments (from `test`)
    segs <- test %>%
      filter(scene_name == scene) %>%
      select(scene_name, reasoning_trace, start_x, start_y, end_x, end_y)

    # Human points
    classified_1 <- classified_subject %>%
      filter(scene_name == scene) %>%
      select(x, y, pupil, gaze_speed, segment_class)

    if (nrow(segs) == 0 || nrow(classified_1) == 0) {
      results_prop_list[[scene]] <- NULL
      next
    }

    rows <- vector("list", nrow(segs))
    keep <- 0

    for (i in seq_len(nrow(segs))) {

      ax <- segs$start_x[i]; ay <- segs$start_y[i]
      bx <- segs$end_x[i];   by <- segs$end_y[i]

      if (!is.finite(ax) || !is.finite(ay) || !is.finite(bx) || !is.finite(by)) next

      # 1) y-range filter only
      y_lo <- min(ay, by)
      y_hi <- max(ay, by)

      cand <- classified_1 %>%
        filter(y >= y_lo, y <= y_hi)

      if (nrow(cand) == 0) next

      # 2) perpendicular distance filter
      d <- point_to_segment_dist(cand$x, cand$y, ax, ay, bx, by)
      nearby <- cand[d <= dist_thresh, , drop = FALSE]

      if (nrow(nearby) == 0) next

      nearby_saccade <- nearby %>% filter(segment_class == "saccade")

      keep <- keep + 1
      rows[[keep]] <- tibble(
        scene_name = scene,
        reasoning_trace = segs$reasoning_trace[i],

        seg_start_x = ax, seg_start_y = ay,
        seg_end_x   = bx, seg_end_y   = by,

        n_within_1 = nrow(nearby),
        mean_pupil = mean(nearby$pupil, na.rm = TRUE),

        mean_gaze_speed_all =
          mean(nearby$gaze_speed, na.rm = TRUE),

        mean_gaze_speed_saccade =
          mean(nearby_saccade$gaze_speed, na.rm = TRUE)
      )
    }

    if (keep == 0) {
      results_prop_list[[scene]] <- NULL
      next
    }

    results_prop_list[[scene]] <- bind_rows(rows[seq_len(keep)])
  }

  results_prop_by_subject[[subject_id]] <- results_prop_list
  message("Completed subject: ", subject_id, "\n")
}

```






## Gaze speed

Calculates each participant’s mean gaze speed during abstraction and simulation, then reports the within-subject difference.


```{r}
results_prop_by_subject_trimmed <- purrr::map(
  results_prop_by_subject,
  function(scene_list) {
    purrr::map(
      scene_list,
      function(df) {
        if (is.null(df) || nrow(df) == 0) return(df)

        # only drop columns if they exist
        dplyr::select(df, -dplyr::any_of(c("X1", "X2")))
      }
    )
  }
)

gaze_speed_by_subject <- purrr::imap_dfr(
  results_prop_by_subject_trimmed,
  function(scene_list, participant_id) {

    participant_df <- dplyr::bind_rows(scene_list)
    if (nrow(participant_df) == 0) return(NULL)

    filtered_df <- participant_df %>%
      dplyr::filter(reasoning_trace %in% c("abstraction", "simulation"))

    # skip subjects where the needed column is missing
    if (!"mean_gaze_speed_all" %in% names(filtered_df)) return(NULL)

    condition_summary <- filtered_df %>%
      dplyr::group_by(reasoning_trace) %>%
      dplyr::summarise(
        mean_gaze_speed = mean(mean_gaze_speed_all, na.rm = TRUE),
        .groups = "drop"
      )

    if (!all(c("abstraction", "simulation") %in% condition_summary$reasoning_trace)) {
      return(NULL)
    }

    abstraction_gaze <- condition_summary$mean_gaze_speed[
      condition_summary$reasoning_trace == "abstraction"
    ]
    simulation_gaze <- condition_summary$mean_gaze_speed[
      condition_summary$reasoning_trace == "simulation"
    ]

    tibble::tibble(
      participant_id = participant_id,
      abstraction = abstraction_gaze,
      simulation = simulation_gaze,
      gaze_speed_diff = simulation_gaze - abstraction_gaze
    )
  }
)

knitr::kable(gaze_speed_by_subject, digits = 3)
```

Runs a paired t-test comparing overall gaze speed between abstraction and simulation.

```{r}
library(dplyr)
library(tidyr)

t_test_result <- t.test(
  gaze_speed_by_subject$abstraction,
  gaze_speed_by_subject$simulation,
  paired = TRUE
)

t_test_result
```






## saccade speed


Calculates each participant’s mean saccade-only gaze speed during abstraction and simulation.

```{r}
results_prop_by_subject_trimmed <- purrr::map(
  results_prop_by_subject,
  function(scene_list) {
    purrr::map(
      scene_list,
      function(df) {
        if (is.null(df) || nrow(df) == 0) return(df)

        dplyr::select(df, -dplyr::any_of(c("X1", "X2")))
      }
    )
  }
)

saccade_gaze_speed_by_subject <- purrr::imap_dfr(
  results_prop_by_subject_trimmed,
  function(scene_list, participant_id) {

    participant_df <- dplyr::bind_rows(scene_list)
    if (nrow(participant_df) == 0) return(NULL)

    filtered_df <- participant_df %>%
      dplyr::filter(reasoning_trace %in% c("abstraction", "simulation"))

    if (!"mean_gaze_speed_saccade" %in% names(filtered_df)) return(NULL)

    condition_summary <- filtered_df %>%
      dplyr::group_by(reasoning_trace) %>%
      dplyr::summarise(
        mean_saccade_gaze_speed = mean(mean_gaze_speed_saccade, na.rm = TRUE),
        .groups = "drop"
      )

    if (!all(c("abstraction", "simulation") %in% condition_summary$reasoning_trace)) {
      return(NULL)
    }

    abstraction_saccade_gaze <- condition_summary$mean_saccade_gaze_speed[
      condition_summary$reasoning_trace == "abstraction"
    ]
    simulation_saccade_gaze <- condition_summary$mean_saccade_gaze_speed[
      condition_summary$reasoning_trace == "simulation"
    ]

    tibble::tibble(
      participant_id = participant_id,
      abstraction = abstraction_saccade_gaze,
      simulation = simulation_saccade_gaze,
      saccade_gaze_speed_diff = simulation_saccade_gaze - abstraction_saccade_gaze
    )
  }
)

knitr::kable(saccade_gaze_speed_by_subject, digits = 3)
```

Runs a paired t-test comparing saccade speed between abstraction and simulation.

```{r}
library(dplyr)
library(tidyr)

t_test_result <- t.test(
  saccade_gaze_speed_by_subject$abstraction,
  saccade_gaze_speed_by_subject$simulation,
  paired = TRUE
)

t_test_result
```

Visualizes within-subject differences for overall gaze speed and saccade speed using violin plots with mean and standard-error markers.

```{r}
library(dplyr)
library(ggplot2)

# Compute paired differences
plot_df <- bind_rows(
  gaze_speed_by_subject %>%
    transmute(
      measure = "gaze speed",
      diff = abstraction - simulation
    ),
  saccade_gaze_speed_by_subject %>%
    transmute(
      measure = "saccade speed",
      diff = abstraction - simulation
    )
)

# Summary stats for optional overlays
summary_df <- plot_df %>%
  group_by(measure) %>%
  summarise(
    mean_diff = mean(diff, na.rm = TRUE),
    se = sd(diff, na.rm = TRUE) / sqrt(sum(!is.na(diff))),
    .groups = "drop"
  )

p <- ggplot(plot_df, aes(x = measure, y = diff, fill = measure)) +
  geom_violin(trim = FALSE, alpha = 0.8, color = NA) +
  geom_jitter(width = 0.08, size = 1.8, alpha = 0.5, color = "black") +
  geom_point(
    data = summary_df,
    aes(x = measure, y = mean_diff),
    inherit.aes = FALSE,
    size = 2.5
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_errorbar(
    data = summary_df,
    aes(
      x = measure,
      ymin = mean_diff - se,
      ymax = mean_diff + se
    ),
    inherit.aes = FALSE,
    width = 0.12,
    size = 0.7
  ) +
  scale_fill_manual(values = c("#40BCD8", "coral")) +
  labs(
    x = NULL,
    y = "within-subject difference\n(abstraction - simulation)"
  ) +
  theme_classic(base_size = 14) +
  theme(
    axis.text.x = element_text(size = 16),
    legend.position = "none"
  )

p
```







# Pupil


Calculates each participant’s mean log pupil size during abstraction and simulation, along with the simulation-minus-abstraction difference.

```{r}
results_prop_by_subject_clean <- purrr::map(
  results_prop_by_subject,
  function(scene_list) {
    purrr::map(
      scene_list,
      function(df) {
        if (is.null(df) || nrow(df) == 0) return(df)
        df[, seq_len(ncol(df) - 2)]
      }
    )
  }
)

pupil_wide <- purrr::imap_dfr(
  results_prop_by_subject_clean,
  function(scene_list, subject_id) {

    subject_df <- dplyr::bind_rows(scene_list)
    if (nrow(subject_df) == 0) return(NULL)

    df <- subject_df %>%
      dplyr::filter(reasoning_trace %in% c("abstraction", "simulation"))

    trace_summary <- df %>%
      dplyr::group_by(reasoning_trace) %>%
      dplyr::summarise(
        mean_mean_pupil = mean(mean_pupil, na.rm = TRUE),
        n          = dplyr::n(),
        sd_pupil   = ifelse(n > 1, sd(mean_pupil, na.rm = TRUE), 0),
        .groups = "drop"
      )

    # must have both conditions
    if (!all(c("abstraction", "simulation") %in% trace_summary$reasoning_trace)) {
      return(NULL)
    }

    mean_abs <- trace_summary$mean_mean_pupil[
      trace_summary$reasoning_trace == "abstraction"
    ]
    mean_sim <- trace_summary$mean_mean_pupil[
      trace_summary$reasoning_trace == "simulation"
    ]

    sd_abs <- trace_summary$sd_pupil[
      trace_summary$reasoning_trace == "abstraction"
    ]
    sd_sim <- trace_summary$sd_pupil[
      trace_summary$reasoning_trace == "simulation"
    ]

    n_abs <- trace_summary$n[
      trace_summary$reasoning_trace == "abstraction"
    ]
    n_sim <- trace_summary$n[
      trace_summary$reasoning_trace == "simulation"
    ]

    se_diff <- sqrt((sd_sim^2 / n_sim) + (sd_abs^2 / n_abs))

    tibble::tibble(
      subject_id = subject_id,
      abstraction = mean_abs,
      simulation  = mean_sim,
      pupil_diff       = mean_sim - mean_abs,
      se_pupil_diff    = se_diff
    )
  }
)

knitr::kable(pupil_wide, digits = 3)
```


Runs a paired t-test testing whether pupil size differs between abstraction and simulation.

```{r}
library(dplyr)
library(tidyr)

t_test_result <- t.test(
  pupil_wide$abstraction,
  pupil_wide$simulation,
  paired = TRUE
)

t_test_result

```



Visualizes the within-subject pupil-size difference between simulation and abstraction using a violin plot with mean and standard-error markers.


```{r, fig.height=4, fig.width=4}
library(dplyr)
library(ggplot2)

# Compute paired differences
plot_df <- pupil_wide %>%
  transmute(
    measure = "pupil size",
    diff = simulation - abstraction
  )

# Summary stats for optional overlays
summary_df <- plot_df %>%
  group_by(measure) %>%
  summarise(
    mean_diff = mean(diff, na.rm = TRUE),
    se = sd(diff, na.rm = TRUE) / sqrt(sum(!is.na(diff))),
    .groups = "drop"
  )

p <- ggplot(plot_df, aes(x = measure, y = diff, fill = measure)) +
  geom_violin(trim = FALSE, alpha = 0.8, color = NA) +
  geom_jitter(width = 0.08, size = 1.8, alpha = 0.5, color = "black") +
  geom_point(
    data = summary_df,
    aes(x = measure, y = mean_diff),
    inherit.aes = FALSE,
    size = 2.5
  ) +
  geom_errorbar(
    data = summary_df,
    aes(
      x = measure,
      ymin = mean_diff - se,
      ymax = mean_diff + se
    ),
    inherit.aes = FALSE,
    width = 0.12,
    size = 0.7
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  scale_fill_manual(values = c("pupil size" = "honeydew3")) +
  labs(
    x = NULL,
    y = "within-subject difference\n(simulation - abstraction)"
  ) +
  theme_classic(base_size = 14) +
  theme(
    axis.text.x = element_text(size = 16),
    legend.position = "none"
  )

p
```








