---
title: "straight_path_analysis_paired_t"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Loads required packages and reads classified_trial_data.csv, a preprocessed dataset generated from notebooks/empirical_gaze_analysis.ipynb. 

```{r}
library(readr)
library(dplyr)
library(purrr)
library(tidyr)
library(tibble)
library(ggplot2)
classified_trial_data <- read_csv(".../classified_trial_data.csv")

drop_scenes <- c(
  "med_nocol_nosp_1",
  "high_yescol_yessp_1",
  "high_nocol_nosp_2",
  "high_nocol_yessp_1",
  "high_yescol_yessp_4",
  "high_nocol_yessp_2",
  "high_yescol_yessp_3",
  "low_yescol_nosp_2"
)

classified_trial_data <- classified_trial_data %>% filter(!scene_name %in% drop_scenes)

classified_trial_data <- classified_trial_data %>%
  dplyr::mutate(
    normalized_time = as.numeric(normalized_time),
    x = as.numeric(x),
    y = as.numeric(y)
  ) %>%
  dplyr::group_by(scene_name, subject_id) %>%
  dplyr::arrange(normalized_time, .by_group = TRUE) %>%
  dplyr::mutate(
    gaze_speed = sqrt((x - dplyr::lag(x))^2 + (y - dplyr::lag(y))^2),
    gaze_speed = dplyr::if_else(
      dplyr::row_number() == 1,
      dplyr::lead(gaze_speed),
      gaze_speed
    )
  ) %>%
  dplyr::ungroup()
classified_trial_data$pupil <- log(classified_trial_data$pupil)

dfs <- classified_trial_data %>%
  dplyr::group_by(subject_id) %>%
  dplyr::group_split(.keep = TRUE)

subject_names <- classified_trial_data %>%
  dplyr::group_by(subject_id) %>%
  dplyr::group_keys() %>%
  dplyr::pull(subject_id)

base::names(dfs) <- subject_names
base::length(dfs)      
base::names(dfs)   
```



For each subject, calculates condition-level summaries for straight-path and no-straight-path trials: mean gaze speed, proportion of saccades, and proportion of smooth pursuit. The results are reshaped into a wide table so each subject has paired values for both path conditions.


```{r}
library(dplyr)
library(purrr)
library(tidyr)

if (is.null(names(dfs))) {
  names(dfs) <- paste0("df", seq_along(dfs))
}

result_table <- imap_dfr(dfs, ~{
  .x %>%
    group_by(path_condition) %>%
    summarise(
      mean_gaze_speed = mean(gaze_speed, na.rm = TRUE),
      prop_saccade = mean(segment_class == "saccade", na.rm = TRUE),
      prop_smooth_pursuit = mean(segment_class == "smooth_pursuit", na.rm = TRUE),
      .groups = "drop"
    ) %>%
    filter(path_condition %in% c("no_straight_path", "straight_path")) %>%
    mutate(dataframe_name = .y) %>%
    pivot_wider(
      names_from = path_condition,
      values_from = c(mean_gaze_speed, prop_saccade, prop_smooth_pursuit)
    )
})

```


Creates a plotting dataset of within-subject condition differences, computes mean differences and standard errors, and defines a reusable violin-plot function. It generates three plots showing subject-level differences for gaze speed, saccade proportion, and smooth-pursuit proportion, then combines them into one labeled figure.


```{r, fig.height=4, fig.width=12}
library(dplyr)
library(ggplot2)
library(patchwork)

# Build plotting dataframe
plot_df <- result_table %>%
  transmute(
    `gaze speed` = mean_gaze_speed_straight_path - mean_gaze_speed_no_straight_path,
    `proportion of saccade` = prop_saccade_straight_path - prop_saccade_no_straight_path,
    `proportion of smooth pursuit` = prop_smooth_pursuit_no_straight_path - prop_smooth_pursuit_straight_path
  ) %>%
  tidyr::pivot_longer(
    cols = everything(),
    names_to = "measure",
    values_to = "diff"
  )

# Summary stats
summary_df <- plot_df %>%
  group_by(measure) %>%
  summarise(
    mean_diff = mean(diff, na.rm = TRUE),
    se = sd(diff, na.rm = TRUE) / sqrt(sum(!is.na(diff))),
    .groups = "drop"
  )

# Reusable plotting function
make_violin <- function(measure_name, fill_color, ylab_text) {
  ggplot(
    dplyr::filter(plot_df, measure == measure_name),
    aes(x = measure, y = diff, fill = measure)
  ) +
    geom_violin(trim = FALSE, alpha = 0.8, color = NA) +
    geom_jitter(width = 0.08, size = 1.8, alpha = 0.5, color = "black") +
    geom_point(
      data = dplyr::filter(summary_df, measure == measure_name),
      aes(x = measure, y = mean_diff),
      inherit.aes = FALSE,
      size = 2.5
    ) +
    geom_errorbar(
      data = dplyr::filter(summary_df, measure == measure_name),
      aes(
        x = measure,
        ymin = mean_diff - se,
        ymax = mean_diff + se
      ),
      inherit.aes = FALSE,
      width = 0.12,
      size = 0.7
    ) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
    scale_fill_manual(values = setNames(fill_color, measure_name)) +
    labs(
      x = NULL,
      y = ylab_text
    ) +
    theme_classic(base_size = 14) +
    theme(
      axis.text.x = element_text(size = 16),
      legend.position = "none"
    )
}

p1 <- make_violin(
  measure_name = "gaze speed",
  fill_color = "#40BCD8",
  ylab_text = "within-subject difference\n(straight path - no straight path)"
)

p2 <- make_violin(
  measure_name = "proportion of saccade",
  fill_color = "coral",
  ylab_text = "within-subject difference\n(straight path - no straight path)"
)

p3 <- make_violin(
  measure_name = "proportion of smooth pursuit",
  fill_color = "lightgreen",
  ylab_text = "within-subject difference\n(no straight path - straight path)"
)

p <- p1 + p2 + p3 +
  patchwork::plot_annotation(tag_levels = list(c("(a)", "(b)", "(c)")))
```






Runs paired t-tests comparing no-straight-path versus straight-path conditions for mean gaze speed, saccade proportion, and smooth-pursuit proportion. The test outputs are printed to evaluate whether the within-subject differences are statistically reliable.

```{r}
t_mean_gaze_speed <- t.test(
  result_table$mean_gaze_speed_no_straight_path,
  result_table$mean_gaze_speed_straight_path,
  paired = TRUE
)

t_prop_saccade <- t.test(
  result_table$prop_saccade_no_straight_path,
  result_table$prop_saccade_straight_path,
  paired = TRUE
)

t_prop_smooth_pursuit <- t.test(
  result_table$prop_smooth_pursuit_no_straight_path,
  result_table$prop_smooth_pursuit_straight_path,
  paired = TRUE
)

t_mean_gaze_speed
t_prop_saccade
t_prop_smooth_pursuit
```











